---
title: "Finding Suitable Locations for Marine Aquaculture"
description: "Cleaning and reclassifying geospatial vector and raster data to pinpoint optimal zones."
author:
  - name: Victoria Cutler
    url: https://victoriacutler.github.io # can also add orchid id under here
    affiliation: MEDS 
    affiliation-url: https://ucsb-meds.github.io
date: 2023-07-20
categories: [GEOSPATIAL DATA ANALYSIS, MAPPING, RASTERS, SHAPEFILES, RECLASSIFYING, MASKING]
citation:
  url: https://victoriacutler.github.io/posts/suitable-habitats/
bibliography: references.bib
format:
  html:
    code-fold: false
draft: false # "true" will mean this is a draft post so it wont show up on my site
---

## Why is Marine Aquaculture Important?

Marine aquaculture holds tremendous promise in addressing the global demand for sustainable protein sources, surpassing conventional land-based meat production. Recent research conducted by Gentry et al. has shed light on the vast potential of marine aquaculture worldwide, considering various crucial factors like ship traffic, dissolved oxygen, and bottom depth. [@gentry2017]

## Finding Suitable Locations for Aquaculture 

In this post, we embark on a journey to identify the most suitable Exclusive Economic Zones (EEZ) along the West Coast of the US for cultivating several oyster species. Our objective is to pinpoint areas where the ideal conditions for oyster growth, such as sea surface temperature and depth, converge, setting the stage for thriving oyster aquaculture ventures.

## Metadata

#### Sea Surface Temperature

We will use average annual sea surface temperature (SST) from the years 2008 to 2012 to characterize the average sea surface temperature within the region. The data we are working with was originally generated from [NOAA's 5km Daily Global Satellite Sea Surface Temperature Anomaly v3.1](https://coralreefwatch.noaa.gov/product/5km/index_5km_ssta.php).

#### Bathymetry

To characterize the depth of the ocean we will use the [General Bathymetric Chart of the Oceans (GEBCO)](https://www.gebco.net/data_and_products/gridded_bathymetry_data/#area).[^1]

[^1]: GEBCO Compilation Group (2022) GEBCO_2022 Grid (<doi:10.5285/e0f0bb80-ab44-2739-e053-6c86abc0289c>).

#### Exclusive Economic Zones

We will be designating maritime boundaries using Exclusive Economic Zones off of the west coast of US from [Marineregions.org](https://www.marineregions.org/eez.php).

## The Code

### Loading the Libraries

```{r warning = FALSE, message = FALSE}
# loading necessary libraries
library(here)
library(sf)
library(terra)
library(raster)
library(dplyr)
library(rpart)
library(rpart.plot)
library(tmap)
```

### Loading and Cleaning Up the Data

Here I read in all the spatial data: west coast zone data, yearly sea surface temperature data, and approximate depth data. I then clean the data by (1) checking and aligning coordinate reference systems, map projections, spatial extent, and spatial resolutions (so that our data are correctly aligned in space) and (2) converting data types and units (for dataset compatibility and ease of future calculations).

**Notes on Methodology:** You'll see that I use the nearest neighbors approach to resample the depth data to match the resolution of the sea surface temperature data, which is of lower resolution. Downscaling like this is a common practice with geospatial datasets since the nearest neighbors approach preserves the original data by assigning the closest pixel in the higher resolution dataset (in this case the closest depth data point) to the newly created lower resolution pixel. I chose this method primarily due to its computational efficiency. Since depth data is continuous, for more accurate analysis something like bilinear or cubic interpolation may be better here.

```{r warning = FALSE, message = FALSE}
# set working directory to here
setwd(here())

# read in the shapefile of the maritime boundaries of the west coast region
wc_regions <- st_read(here("posts", "suitable-habitats", "data", "wc_regions_clean.shp"))

# read in the yearly sea surface temperature (SST) raster layers
average_annual_sst_2008 <- rast(here("posts", "suitable-habitats", "data", "average_annual_sst_2008.tif"))
average_annual_sst_2009 <- rast(here("posts", "suitable-habitats", "data", "average_annual_sst_2009.tif"))
average_annual_sst_2010 <- rast(here("posts", "suitable-habitats", "data", "average_annual_sst_2010.tif"))
average_annual_sst_2011 <- rast(here("posts", "suitable-habitats", "data", "average_annual_sst_2011.tif"))
average_annual_sst_2012 <- rast(here("posts", "suitable-habitats", "data", "average_annual_sst_2012.tif"))

# combine rasters into a stack
avg_annual_sst_stack <- stack(c(average_annual_sst_2008, average_annual_sst_2009, average_annual_sst_2010, average_annual_sst_2011, average_annual_sst_2012))

# read in the bathymetry (aka depth) raster
depth <- rast(here("posts", "suitable-habitats", "data", "depth.tif"))

# check the coordinate reference systems (CRSs)
cat(crs(wc_regions))
cat(as.character(crs(avg_annual_sst_stack)))
cat(crs(depth))

  # converting the STT stack back to a SpatRaster to use terra functions, like to change the crs
avg_annual_sst_stack <- rast(avg_annual_sst_stack)

  # converting the STT stack SpatRast to the matching crs
crs(avg_annual_sst_stack) = "EPSG:4326"

# find the mean SST then convert to Celcius
average_sst_degC <- mean(avg_annual_sst_stack, na.rm = TRUE) - 273.15

# crop the depth raster to match the extent of the SST raster
depth_newextent <- crop(depth, average_sst_degC) # you can see that the extents are now the same

# resample the depth raster such that the resolution is the same as the resolution in the SST raster
depth_newres <- resample(x = depth_newextent, y = average_sst_degC, method = "near")

# checking that the dimensions, resolutions, extent, and crs match
average_sst_degC
depth_newres
  # doing this by checking to see that the two rasters can be stacked
stack_test <- rast(c(average_sst_degC, depth_newres))
stack_test # stackable
```

### Using Reclassification Matrices to Assess Oyster Ecosystem Suitability 

In this process, I conduct a reclassification of both the sea surface temperature raster and depth raster. Areas with suitable depth or temperature for oyster aquaculture are assigned a value of '1', while unsuitable areas are marked as 'NA'. By subsequently employing the `lapply()` function, I perform a multiplication operation on the reclassified rasters. The outcome is a final raster presenting a clear distinction between '1s' indicating suitable habitat and 'NAs' representing areas unsuitable for oyster cultivation, providing a comprehensive depiction of overall suitability.

```{r}
# reclassifying the SST raster for oyster suitability (11-30°C)

  # first creating a reclassification matrix
rcl_sst <- matrix(c(-Inf, 11, NA,
                    11, 30, 1,
                 30, Inf, NA), ncol = 3, byrow = TRUE)
  # then reclassify the matrix
sst_suitable <- classify(average_sst_degC, rcl = rcl_sst)

# reclassifying the depth raster for oyster suitability (-70-0)

  # first creating a reclassification matrix
rcl_depth <- matrix(c(-Inf, -70, NA,
                    -70, 0, 1,
                 0, Inf, NA), ncol = 3, byrow = TRUE)
  # then reclassify the matrix
depth_suitable <- classify(depth_newres, rcl = rcl_depth)

# finding locations where depth and sst are suitable for oysters
  # making a raster multiplication function
mult_fun = function(rast1, rast2){
  rast1 * rast2
}

  # creating the raster that has NAs where both are not suitable, and 1s where depth and sst are suitable
depth_and_stt_suitable <- lapp(c(depth_suitable, sst_suitable), fun = mult_fun)
```

### Scaling to Zones of Interest (Ecological Economic Zones)

```{r}

# data exploration - are these rasters of the same extent/crs etc? yes!
wc_regions
depth_and_stt_suitable

  # first, make wc_regions into a raster
wc_regions_simp <- wc_regions |> 
  select("rgn")

wc_regions_rast <- rasterize(x = wc_regions_simp, y = depth_and_stt_suitable, field = "rgn")

# finding the suitable cells within each region
  # first mask the region raster with the suitable cells (since not suitable cells are NA, our region raster will only have values where suitable)
wc_regions_mask <- mask(wc_regions_rast, depth_and_stt_suitable)

  # lastly, displaying our suitable cells within each region that are suitable
tm_shape(wc_regions_mask) +
  tm_raster()

# finding the area of the suitable grid cells of each region
suitable_area_byregion <- zonal(cellSize(wc_regions_mask, unit="m"), wc_regions_mask, fun = "sum") |> 
  mutate(area = area / (1000^2)) |> 
  rename(suitable_area_km2 = area)
suitable_area_byregion 

# finding the area of the grid cells for each region (even though we have this already reported in the original datafile, we're using the same method we used to calculate our suitable area to make sure percentages are more likely proportional, since there are slight differences in this area approximation method). 
total_area_byregion <- zonal(cellSize(wc_regions_rast, unit="m"), wc_regions_rast, fun = "sum") |> 
  mutate(area = area / (1000^2)) |> 
  rename(total_area_km2 = area)
total_area_byregion 

# find the percent of suitable area per area per region
percent_area_df <- inner_join(total_area_byregion, suitable_area_byregion, by = "rgn") |> mutate(percent_suitable = (suitable_area_km2/total_area_km2) * 100)

# source code: https://stackoverflow.com/questions/68979884/how-to-loop-through-a-dataframe-and-get-value-of-corresponding-column
for(i in 1:dim(percent_area_df)[1]) {
  writeLines((paste0(percent_area_df$rgn[i], " has an area suitable for oysters of " , round(percent_area_df$suitable_area_km2[i], 2), "km2 which amounts to an area that is ", round(percent_area_df$percent_suitable[i],2), "% suitable for oysters.","\n")))
}

  # test: look at percentages using the given areas to see if theyre similar:
#percent_area_df_originalareas <- merge(wc_regions, suitable_area_byregion, by = "rgn") |> mutate(percent_suitable = (suitable_area_km2/area_km2) * 100)
```

### Visualizing Suitable Ecological Economic Zones

```{r}
# first joining the the area_df back to the region sf object for plotting
region_stats_join <- merge(wc_regions, percent_area_df)

# switching to tmap view mode to add a built in basemap
tmap_mode("view")

# total suitable area by region
  # create legend title
#legend_title_area = expression("Suitable Area for Oysters (km"^2*")")
legend_title_area = expression("Suitable Area for Oysters (km^2)")
area_suitable_map <- tm_shape(region_stats_join) +
  tm_fill(col = "suitable_area_km2",
          title = legend_title_area,
          style = "pretty",
          palette = "Blues") +
  tm_borders() +
  tm_legend(legend.outside = TRUE) +
  #tm_layout(main.title = "West Coast Regions Suitable for Oysters",
            #main.title.size = 1)  +
  tm_text("rgn", size = 0.3) +
  tm_basemap(server = "OpenTopoMap")

# percent suitable area by region
percent_suitable_map <- tm_shape(region_stats_join) +
  tm_fill(col = "percent_suitable",
          title = "Percent of Region Suitable for Oysters",
          style = "pretty",
          palette = "Blues") +
  tm_borders() +
  tm_legend(legend.outside = TRUE) +
 # tm_layout(main.title = "Percent of West Coast Region Suitable for Oysters",
           # main.title.size = 1) +
  tm_text("rgn", size = 0.3) +
  tm_basemap(server = "OpenTopoMap")

tmap_arrange(area_suitable_map, percent_suitable_map)
```

### Creating a Function to Broaden this Workflow

This sort of analysis can be applicable for many different applications. Not only can this analysis be used to assess aquaculture suitability for other species, not just the oyster, but it can also be used to assess critical habitat needs for any key marine life that depends on certain sea surface temperature and ocean depth.

Here I turn my analysis into a function that takes in the marine species of interest, habitat temperature and depth ranges and returns maps of total suitable area and percent suitable area per exclusive economic zone.

Below I test my function using data from [SeaLifeBase](https://www.sealifebase.ca/search.php) to find suitable habitat for the Bottlenose Dolphin. In reality, however, geolocating Bottlenose Dolphin habitat may depend on additional key factors.

```{r}
# creating a function that takes in depth ranges, temperature ranges, and species name
suitability_map_fun = function(shallowest_depth, deepest_depth, min_temp, max_temp, species_name){
  
  # code sourced from the "find suitable locations" section
rcl_sst <- matrix(c(-Inf, min_temp, NA,
                    min_temp, max_temp, 1,
                 max_temp, Inf, NA), ncol = 3, byrow = TRUE)
sst_suitable <- classify(average_sst_degC, rcl = rcl_sst)

rcl_depth <- matrix(c(-Inf, -deepest_depth, NA,
                    -deepest_depth, shallowest_depth, 1,
                 shallowest_depth, Inf, NA), ncol = 3, byrow = TRUE)
depth_suitable <- classify(depth_newres, rcl = rcl_depth)

mult_fun = function(rast1, rast2){
  rast1 * rast2
}
depth_and_stt_suitable <- lapp(c(depth_suitable, sst_suitable), fun = mult_fun)

  # code sourced from the "Determine the most suitable EEZ" section
wc_regions_simp <- wc_regions |> 
  select("rgn")
wc_regions_rast <- rasterize(x = wc_regions_simp, y = depth_and_stt_suitable, field = "rgn")
wc_regions_mask <- mask(wc_regions_rast, depth_and_stt_suitable)

suitable_area_byregion <- zonal(cellSize(wc_regions_mask, unit="m"), wc_regions_mask, fun = "sum") |> 
  mutate(area = area / (1000^2)) |> 
  rename(suitable_area_km2 = area)

total_area_byregion <- zonal(cellSize(wc_regions_rast, unit="m"), wc_regions_rast, fun = "sum") |> 
  mutate(area = area / (1000^2)) |> 
  rename(total_area_km2 = area)

percent_area_df <- inner_join(total_area_byregion, suitable_area_byregion, by = "rgn") |> mutate(percent_suitable = (suitable_area_km2/total_area_km2) * 100)

# code sourced from the "Determine the most suitable EEZ" section
region_stats_join <- merge(wc_regions, percent_area_df)

tmap_mode("view")

legend_title_area = paste0("Suitable Area for the ", species_name, " Species (km^2)")
area_suitable_map <- tm_shape(region_stats_join) +
  tm_fill(col = "suitable_area_km2",
          title = legend_title_area,
          style = "pretty",
          palette = "Blues") +
  tm_borders() +
  tm_legend(legend.outside = TRUE) +
  tm_text("rgn", size = 0.3) +
  tm_basemap(server = "OpenTopoMap")

legend_title_percent = paste0("Percent of Region Suitable for the ", species_name, " Species")
percent_suitable_map <- tm_shape(region_stats_join) +
  tm_fill(col = "percent_suitable",
          title = legend_title_percent,
          style = "pretty",
          palette = "Blues") +
  tm_borders() +
  tm_legend(legend.outside = TRUE) +
  tm_text("rgn", size = 0.3) +
  tm_basemap(server = "OpenTopoMap")

tmap_arrange(area_suitable_map, percent_suitable_map)

}

# running the function for the bottlenose dolphin
  # source: https://www.sealifebase.ca/summary/Tursiops-truncatus.html
suitability_map_fun(shallowest_depth = 0, deepest_depth = 1316, min_temp = 6.4, max_temp = 28.9, species_name = "Dolphin")
```
